---
title: "Plotting read coverage of Sendai and Yamanaka"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(here)
library(tidyverse)
library(powerjoin)
library(data.table)
library(synExtra)
library(ggcoverage)
library(ggbio)
library(GenomicRanges)

synapser::synLogin()

theme_set(theme_minimal(base_family = "Helvetica"))

syn <- synDownloader("~/data", .cache = TRUE)
```

```{r}
inputs <- c(
  sample_meta = "syn53061904",
  yamanaka_coverage = "syn58277625",
  sendai_coverage = "syn58277604"
)

input_files <- map(
  inputs, syn
)

sample_meta <- read_csv(input_files[["sample_meta"]])
yamanaka_coverage <- read_csv(input_files[["yamanaka_coverage"]])
sendai_coverage <- read_csv(input_files[["sendai_coverage"]])
```

```{r}
ensembl_gtf <- rtracklayer::import(
  here("index", "Homo_sapiens.GRCh38.110.gtf.gz")
)
```

Select samples to plot.

```{r}
samples_sendai <- sample_meta %>%
  filter(
    (study_name == "Buckberry Nature 2023" &
      treatment %in% c("NtP", "Naïve", "Primed", "Control ESC", "TNT")) |
      source_name == "5iLAF SSEA4- subpopulation UCLA1 hESCs"
  ) %>%
  # Pastor samples are Control ESC
  mutate(
    treatment = if_else(
      source_name == "5iLAF SSEA4- subpopulation UCLA1 hESCs",
      "Control ESC", treatment
    )
  )

```

## Sendai

Define position of Sendai genes.

```{r}
sendai_gene_df <- tribble(
  ~symbol, ~start, ~end,
  "N gene", 1202, 2884,
  "P gene", 2888, 4570,
  "M gene", 4606, 5652,
  "F gene", 5803, 7500,
  "HN gene", 7630, 9353,
  "L gene", 9493, 16219
) %>%
  mutate(
    seqname = "pSeV-idSOX2-Pmut",
    strand = "+",
    type = "gene",
    gene_name = symbol,
    gene_type = "protein_coding"
  )


sendai_gene_glist <- GenomicRanges::makeGRangesListFromDataFrame(
  mutate(
    sendai_gene_df,
    type = "exon"
  ),
  keep.extra.columns = TRUE,
  split.field = "gene_name"
)
```

Make plot of Sendai "gene models".

```{r}
sendai_gene_plot <- ggplot() +
  geom_alignment(
    sendai_gene_glist,
    # aes(
    #   y = ypos
    # ),
    range.geom = "arrow",
    cds.rect.h = 0.1
  )
```

Only keep two replicates per treatment group.

```{r}
sendai_coverage_filtered <- sendai_coverage %>%
  power_left_join(
    samples_sendai,
    by = "sample_accession",
    check = check_specs(
      unmatched_keys_left = "warn",
      duplicate_keys_right = "warn"
    )
  ) %>%
  mutate(
    treatment_grouped = case_when(
      treatment == "Control ESC" ~ paste("Control ESC", study_name),
      TRUE ~ treatment
    )
  ) %>%
  group_by(treatment_grouped) %>%
  mutate(
    rep = factor(
      sample_accession, labels = seq_len(as.character(length(unique(cur_data()$sample_accession))))
    )
  ) %>%
  # Only keep two replicates
  filter(
    rep %in% c("1", "2")
  ) %>%
  ungroup()

```

Make coverage plot of Sendai genome for chosen range that includes the relevant genes.

```{r}
X_RANGE <- c(1050, 16400)

sendai_coverage_plot <- ggplot(
  sendai_coverage_filtered %>%
    mutate(
      score = pmax(score, 1e-8)
    ),
  aes(
    x = pos,
    y = score,
    fill = rep,
    group = sample_accession
  )
) +
  # geom_line() +
  # geom_ribbon(aes(ymax = score), ymin = -Inf, alpha = 0.5, line.width = NA, outline.type = "upper") +
  # geom_step(alpha = 0.6) +
  # geom_line(aes(y = score), alpha = 0.5) +
  geom_ribbon(aes(ymax = score), ymin = -Inf, alpha = 0.6, outline.type = "upper", color = NA) +
  ggh4x::facet_wrap2(~treatment_grouped, ncol = 1, strip.position = "right", axes = "all", remove_labels = "x") +
  geom_text(
    aes(x = X_RANGE[1] + 500, y = 1e2, label = treatment_grouped),
    data = distinct(sendai_coverage_filtered, treatment_grouped),
    inherit.aes = FALSE,
    hjust = 0, vjust = 1,
    size = 3
  ) +
  guides(
    # color = "none",
    # fill = "none"
  ) +
  paletteer::scale_fill_paletteer_d("colorblindr::OkabeIto") +
  # geom_gene(gene_granges) +
  scale_y_log10(
    labels = scales::label_log(),
    breaks = c(1e-3, 1e2),
    limits = c(1e-3, 1e2),
    expand = c(0, 0),
    oob = scales::squish_infinite
  ) +
  # scale_y_log10() +
  labs(
    y = "CPM", color = "Replicate", fill = "Replicate"
  ) +
  theme_classic() +
  theme(
    strip.background = element_blank(),
    strip.text = element_blank(),
    panel.grid.minor.y = element_blank(),
    panel.grid.major.y = element_blank(),
    axis.text.y = element_text(size = rel(0.8)),
    # axis.line.x = element_line(size = 1),
    legend.position = "top"
  )

ggplot2::ggsave(
  "plots/sendai_coverage_plot_raw.pdf",
  sendai_coverage_plot,
  width = 5, height = 4
)
```

Combine coverage plot with gene model plot.

```{r}
sendai_combined <- patchwork::wrap_plots(
  sendai_coverage_plot +
    coord_cartesian(xlim = X_RANGE, expand = FALSE) +
    theme(
      # axis.text.x = element_blank(),
      axis.title.x = element_blank(),
      legend.key.size = unit(8, "pt"),
      legend.title = element_text(size = unit(8, "pt")),
      legend.text = element_text(size = rel(0.8)),
      legend.margin = margin(0, 0, 0, 0)
    ),
  sendai_gene_plot +
    coord_cartesian(clip = "off", xlim = X_RANGE, expand = FALSE) +
    theme(
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank(),
      axis.text.x = element_blank(),
      axis.title.x = element_blank()
    ),
  heights = c(1, 0.15)
)

ggplot2::ggsave(
  "plots/sendai_coverage_tracks.pdf",
  sendai_combined,
  width = 4, height = 3.5
)

sendai_combined <- tracks(
  sendai_coverage_plot,
  sendai_gene_plot,
  ncol = 1,
  heights = c(1, 0.15)
)

xlim(sendai_combined) <- c(1150, 16000)

withr::with_cairo_pdf(
  "plots/sendai_coverage_tracks.pdf",
  width = 5, height = 4,
  {
    print(sendai_combined)
  }
)
```

## Yamanaka

Prepare Yamanaka gene CDS and UTR coordinates.

```{r}
yamanaka_factors <- c("KLF4", "MYC", "POU5F1", "SOX2")

yamanaka_gtf <- ensembl_gtf[
  ensembl_gtf$gene_name %in% yamanaka_factors &
    replace_na(str_detect(ensembl_gtf$tag, "MANE_Select"), FALSE)
]

yamanaka_gtf_edited <- yamanaka_gtf
yamanaka_gtf_edited$type <- recode(
  yamanaka_gtf_edited$type,
  CDS = "cds",
  five_prime_utr = "utr",
  three_prime_utr = "utr"
)
yamanaka_gtf_edited <- yamanaka_gtf_edited[yamanaka_gtf_edited$type %in% c("cds", "utr", "gap")]

yamanaka_regions <- reduce(
  split(yamanaka_gtf, yamanaka_gtf$gene_name)
)
```

Plot each gene's coverage using counts rescaled to [0, 1] range.

```{r}
yamanaka_coverage_rescaled <- yamanaka_coverage %>%
  group_by(sample_accession, gene_name) %>%
  mutate(
    score_norm = scales::rescale(score, to = c(0, 1))
  ) %>%
  ungroup() %>%
  power_left_join(
    samples_sendai,
    by = "sample_accession",
    check = check_specs(
      unmatched_keys_left = "warn",
      duplicate_keys_right = "warn"
    )
  ) %>%
  mutate(
    treatment_grouped = case_when(
      treatment == "Control ESC" ~ paste("Control ESC", str_split_fixed(study_name, " ", 3)[, 1], "et al."),
      TRUE ~ treatment
    )
  ) %>%
  group_by(treatment_grouped) %>%
  mutate(
    rep = factor(
      sample_accession, labels = seq_len(as.character(length(unique(cur_data()$sample_accession))))
    )
  ) %>%
  # Only keep two replicates
  filter(
    rep %in% c("1", "2")
  ) %>%
  ungroup()
```

Pastor et al samples have a strong 3' bias, can't use them here.

```{r}
yamanaka_plots <- imap(
  yamanaka_regions,
  \(region, gene_name) {
    # browser()
    region_str <- paste0(seqnames(region), ":", start(region), "-", end(region))
    gene_plot <- ggplot() +
      geom_alignment(
        yamanaka_gtf_edited[yamanaka_gtf_edited$gene_name == gene_name] %>%
          split(.$gene_name),
        cds.rect.h = 0.05
      )

    utr_regions <- yamanaka_gtf_edited %>%
      as_tibble() %>%
      filter(
        .data$gene_name == gene_name,
        type == "utr"
      )

    bam_coverage_long <- yamanaka_coverage_rescaled %>%
      filter(
        .data$gene_name == gene_name,
        study_name != "Pastor CSC 2016"
      )

    X_RANGE <- c(start(region), end(region))
    X_LEN <- end(region) - start(region)
    X_RANGE <- c(X_RANGE[1] - 0.05 * X_LEN, X_RANGE[2] + 0.05 * X_LEN)
    Y_SCALE <- c(0, 1)

    yamanaka_coverage_plot <- ggplot(
      bam_coverage_long %>%
        mutate(
          score_norm = pmax(score_norm, 1e-8)
        ),
      aes(
        x = pos,
        y = score_norm,
        fill = rep,
        group = sample_accession
      )
    ) +
      geom_rect(
        aes(xmin = start, xmax = end),
        ymin = -Inf, ymax = Inf,
        fill = alpha("grey", 0.4),
        data = utr_regions,
        inherit.aes = FALSE
      ) +
      geom_ribbon(aes(x = pos, ymax = score_norm), ymin = -Inf, alpha = 0.6, outline.type = "upper", color = NA) +
      ggh4x::facet_wrap2(~treatment_grouped, ncol = 1, strip.position = "right", axes = "all", remove_labels = "x") +
      geom_text(
        aes(x = X_RANGE[1] + .02 * X_LEN, y = Y_SCALE[2], label = treatment_grouped),
        data = distinct(bam_coverage_long, treatment_grouped),
        inherit.aes = FALSE,
        hjust = 0, vjust = 1,
        size = 3
      ) +
      paletteer::scale_fill_paletteer_d("colorblindr::OkabeIto") +
      scale_y_continuous(
        breaks = Y_SCALE,
        limits = Y_SCALE,
        expand = c(0, 0),
        oob = scales::squish_infinite
      ) +
      scale_x_continuous(
        labels = scales::label_number(scale_cut = scales::cut_short_scale(), suffix = "b")
      ) +
      labs(
        y = "Rescaled CPM", color = "Replicate", fill = "Replicate"
      ) +
      theme_classic() +
      theme(
        strip.background = element_blank(),
        strip.text = element_blank(),
        panel.grid.minor.y = element_blank(),
        panel.grid.major.y = element_blank(),
        axis.text.y = element_text(size = rel(0.8)),
        legend.position = "top"
      )

    combined <- patchwork::wrap_plots(
      yamanaka_coverage_plot +
        coord_cartesian(xlim = X_RANGE, expand = FALSE) +
        theme(
          axis.title.x = element_blank(),
          axis.text.x = element_blank(),
          legend.key.size = unit(8, "pt"),
          legend.title = element_text(size = unit(8, "pt")),
          legend.text = element_text(size = rel(0.8)),
          legend.margin = margin(0, 0, 0, 0)
        ),
      gene_plot +
        coord_cartesian(clip = "off", xlim = X_RANGE, expand = FALSE) +
        theme(
          panel.grid.major = element_blank(),
          panel.grid.minor = element_blank(),
          axis.text.x = element_blank(),
          axis.title.x = element_blank()
        ),
      heights = c(1, 0.06)
    )

    combined
  }
)

iwalk(
  yamanaka_plots,
  \(plot, gene_name) {
    withr::with_cairo_pdf(
      paste0("plots/yamanaka_", gene_name, "_coverage_tracks_range_norm_lin.pdf"),
      width = 4, height = 3,
      {
        print(plot)
      }
    )
  }
)
```

Combined all four genes into a single plot.

```{r}
# Errors on cluster without dummy device
withr::with_pdf(
  "dummpy.pdf",
  yamanaka_plots_combined <- gridExtra::grid.arrange(
    grobs = map2(
      yamanaka_plots,
      seq_along(yamanaka_plots),
      \(x, i) {
        # browser()
        x[[1]] <- x[[1]] +
          # guides(
          #   fill = if(i == 1) guide_legend() else "none"
          # ) +
          theme(
            axis.title.y = if(i == 1) element_text() else element_blank(),
            axis.text.y = if(i == 1) element_text() else element_blank()
          )
        x <- x & theme(
          plot.margin = margin(0, 0, 8, 0),
          axis.text.x = element_blank(),
          axis.ticks.x = element_blank()
        )
        patchwork:::as_patch(x)
      }
    ),
    nrow = 1
  )
)

withr::with_cairo_pdf(
  "plots/yamanaka_coverage_tracks_rescaled.pdf",
  grid::grid.draw(yamanaka_plots_combined),
  width = 8, height = 3
)


```
